---
title: "Case Study--All Vairables Treated Equally "
author: "Tuul Purevjav"
date: "3/2/2022"
output: word_document
---

```{r}
library(dplyr)
library(tidyverse)
library(GGally)
library(car)
```

My analysis is treating all variables as equals (without giving special treatment to Statin).

## Summary Statistics and Graphics

```{r}
prevend <- read.csv("Prevend_Sample.csv")
```

Check for NA's
```{r}
sum(is.na(prevend))
```

**Summary Statistics**
```{r}
summary(prevend)
```

Handling NA's:
```{r}
#replacing negative values in the 4 relevant columns with NA
prevend$VAT <- replace(prevend$VAT, prevend$VAT < 0, NA)
prevend$Smoking <- replace(prevend$Smoking, prevend$Smoking < 0, NA)
prevend$HDL <- replace(prevend$HDL, prevend$HDL < 0, NA)
prevend$FRS <- replace(prevend$FRS, prevend$FRS < 0, NA)

#count the number of rows that have an NA value present
sum(!complete.cases(prevend))
```

```{r}
# This removes only 2 NA observations on my machine, not 42. Not sure why.
prevend <-na.omit(prevend)
dim(prevend)
```

```{r}
# remove irrelevant variable `Casenr` from the data frame.
prevend<- prevend[, -1]
```

**Analysis**

I tried to plot a matrix of scatterplots with RFFT vs all variable, it was impossible to see any that made sense. So I removed categorical variables (since they're meaningless anyway) and created a plot of matrix only with RFFT and quantitative variables in the effort to get some sort of visual of their relationship.
```{r}
# create new dataframe dummy variables removed
prevend.no.categ <- prevend[,-c(2,3,4,7,8,9,10,16)]
ggpairs(prevend.no.categ)
```

```{r}
# remove df prevend.no.categ
rm(prevend.no.categ)
```

Check correlations for RFFT with all predictors:
```{r}
# create a matrix of correlations
(cor <- round(cor(prevend[,-5]),2))

# correlation between RFFT and other predictors
round(cor(prevend$RFFT,prevend[,-5]),2)
```
Large correlations with RFFT: 
  Age  -0.53, 
  Education  0.57, 
  FRS  -0.48
Large correlations in general: 
  Age/FRS 0.81, 
  Hypertension/MAP  0.54, 
  SBP/FRS 0.66, 
  DBP/SBP 0.70, 
  MAP/SBP 0.91, 
  MAP/DBP 0.89,
  FRS/Hypertension  0.63,
  MAP/FRS 0.59


**Starting with fitting the full model**

```{r}
m.full <- lm(RFFT ~ . ,data=prevend)
summary(m.full)
```
Taken together, the predictors have Adjusted R-squared 0.4604, which is not that great.  Any of them whose p-value is not close to 0 might be OK to remove. Overall F-statistic is 22.16, which is big and overall p-value is < 2.2e-16f, which is small. 
We have a discrepancy between small R-squared and small overall p-value. This may suggests some of the predictors could be strongly correlated (collinearity problem), which we already know they are (i.e., SBP, DBP and MAP are all blood pressure measurements).
Looking at the variable-added-last test (P-value column), Age, Gender, Education, Smoking, Map, and Statin seem essential b/c their small p-values.

Check the residual plots for the full model:
```{r}
par(mfrow=c(1,3))
plot(m.full,c(1,2,4))
```
Residuals vs Fitted values: appears to have some fanning and might suggest non-constant variance.
QQPlot: looks very straight, suggests that residuals are normal.
Cook's distance: check obs 153, 375, and 481.


Check plots of the residuals against each of quantitative predictors to examine the linearity of the model, mean zero and constant variance:
```{r echo=FALSE, fig.height=12, fig.width=6}
rsd <- resid(m.full)
par(mfrow=c(10,2))
par(mar=c(3.8, 4, 2, 0.5))
plot(prevend$Age,rsd,xlab="Age",ylab="residual")
abline(a=0,b=0)
plot(prevend$Gender,rsd,xlab="Gender",ylab="residual")
abline(a=0,b=0)
plot(prevend$Ethnicity,rsd,xlab="Ethnicity",ylab="residual")
abline(a=0,b=0)
plot(prevend$Education,rsd,xlab="Education",ylab="residual")
abline(a=0,b=0)
plot(prevend$VAT,rsd,xlab="VAT",ylab="residual")
abline(a=0,b=0)
plot(prevend$CVD,rsd,xlab="CVD",ylab="residual")
abline(a=0,b=0)
plot(prevend$DM,rsd,xlab="DM",ylab="residual")
abline(a=0,b=0)
plot(prevend$Smoking,rsd,xlab="Smoking",ylab="residual")
abline(a=0,b=0)
plot(prevend$Hypertension,rsd,xlab="Hypertension",ylab="residual")
abline(a=0,b=0)
plot(prevend$BMI,rsd,xlab="BMI",ylab="residual")
abline(a=0,b=0)
plot(prevend$SBP,rsd,xlab="SBP",ylab="residual")
abline(a=0,b=0)
plot(prevend$DBP,rsd,xlab="DBP",ylab="residual")
abline(a=0,b=0)
plot(prevend$MAP,rsd,xlab="MAP",ylab="residual")
abline(a=0,b=0)
plot(prevend$eGFR,rsd,xlab="eGFR",ylab="residual")
abline(a=0,b=0)
plot(prevend$Albuminuria,rsd,xlab="Albuminuria",ylab="residual")
abline(a=0,b=0)
plot(prevend$Chol,rsd,xlab="Chol",ylab="residual")
abline(a=0,b=0)
plot(prevend$HDL,rsd,xlab="HDL",ylab="residual")
abline(a=0,b=0)
plot(prevend$Statin,rsd,xlab="Statin",ylab="residual")
abline(a=0,b=0)
plot(prevend$FRS,rsd,xlab="FRS",ylab="residual")
abline(a=0,b=0)
```
Age looks good and VAT appears to have increasing variance to the right. BMI has cluttering on the left. Valerie pointed out that the distribution of Ethnicity was off.

Check the distribution of Ethnicity
```{r}
plot(density(prevend$Ethnicity))
```
Like Valerie said, the plot shows most of the individuals were Western European (0 = Western European). We should remove it from the study.

Check a density estimate of RFFT.
```{r}
plot(density(prevend$RFFT))
```
The symmetry is not great but reasonable.

Check the plot of observed responses against fitted responses:
```{r}
ft <- fitted(m.full)
plot(ft,prevend$RFFT,xlab="fitted",ylab="RFFT",main="RFFT vs Fitted")
abline(a=0,b=1,col="red")
```
The straight line fit looks reasonable, and there could be some fanning out a little bit, which may suggest non-constant variance assumption not met? I could try transformation.

**Let's check VIF.**

Check collinearity using variance inflation factors (VIF):
```{r}
vif(m.full)
```
Age and DBP are larger than 5. SBP, MAP and FRS are larger than 10, which is not surprising. This suggests we are dealing with collinearity.

**Let's consider transformation.**

```{r}
boxCox(m.full)
```
The closest “rounded” value for λ is 0.5, which corresponds to the square root transformation.

Let's try transforming the response and the predictors simultaneously:
```{r}
sqrtRFFT <- sqrt(prevend$RFFT)
sqrtAge <- sqrt(prevend$Age)
sqrtGender <- sqrt(prevend$Gender)
sqrtEthnicity <- sqrt(prevend$Ethnicity)
sqrtEducation <- sqrt(prevend$Education)
sqrtVAT <- sqrt(prevend$VAT)
sqrtCVD <- sqrt(prevend$CVD)
sqrtDM <- sqrt(prevend$DM)
sqrtSmoking <- sqrt(prevend$Smoking)
sqrtHypertension <- sqrt(prevend$Hypertension)
sqrtBMI <- sqrt(prevend$BMI)
sqrtSBP <- sqrt(prevend$SBP)
sqrtDBP <- sqrt(prevend$DBP)
sqrtMAP <- sqrt(prevend$MAP)
sqrteGFR <- sqrt(prevend$eGFR)
sqrtAlbuminuria <- sqrt(prevend$Albuminuria)
sqrtChol <- sqrt(prevend$Chol)
sqrtHDL <- sqrt(prevend$HDL)
sqrtStatin <- sqrt(prevend$Statin)
sqrtFRS <- sqrt(prevend$FRS)
sqrt.prevend <- data.frame(sqrtRFFT,sqrtAge,sqrtGender,sqrtEthnicity,sqrtEducation,sqrtVAT,sqrtCVD,
                                sqrtDM,sqrtSmoking,sqrtHypertension,sqrtBMI,sqrtSBP,
                                sqrtDBP,sqrtMAP,sqrteGFR,sqrtAlbuminuria,sqrtChol,sqrtHDL,sqrtStatin,sqrtFRS)
m.sqrt.full<- lm(sqrt.prevend$sqrtRFFT ~ ., data=sqrt.prevend)
summary(m.sqrt.full)
```

```{r}
summary(m.sqrt.full)
```
The Adjusted R-squared increased from 0.4604 to 0.4648. Not a big improvement.

Let's compare the plot of observed values against fitted values for each of the models:
```{r}
par(mfrow=c(2,2))
plot(ft,prevend$RFFT,xlab="fitted",ylab="RFFT",main="RFFT vs Fitted")
abline(a=0,b=1,col="red")

plot(m.sqrt.full$fitted.values,sqrtRFFT,xlab="Fitted vs sqrt(Y)")
abline(a=0,b=1,col="red")
```
Sqrt model fits a little better, but I decided not to consider transformation. The original model is probably reasonably good enough to proceed.

**Let's check AIC and BIC.**

Backward AIC:
```{r}
backAIC <- step(m.full,direction="backward", data=prevend)
```
AIC backward result: RFFT ~ Age + Ethnicity + Education + VAT + CVD + Smoking + Hypertension + 
    SBP + DBP + MAP + eGFR + Chol + Statin

Backward BIC
```{r}
n <- nrow(prevend)
backBIC <- step(m.full, direction="backward", data=prevend, k=log(n))
```
Backward BIC result: RFFT ~ Age + Education + Smoking

Forward AIC:
```{r}
m.full.intercept <- lm(RFFT~1, data=prevend)
forwardAIC <- step(m.full.intercept, scope=list(lower= ~ 1, upper= m.full),
direction="forward", data=prevend)
```
Forward AIC result: RFFT ~ Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + eGFR + BMI

Forward BIC:
```{r}
forwardBIC <- step(m.full.intercept, scope=list(lower= ~ 1, upper= m.full), direction="forward",
data=prevend, k=log(n))
```
Forward BIC Result: RFFT ~ Education + Age + Smoking

Compare summaries of all selected models:
```{r}
# backward AIC
m.backAIC <- lm(RFFT ~ Age + Ethnicity + Education + VAT + CVD + Smoking + Hypertension +SBP + DBP + MAP + eGFR + Chol + Statin)
summary(m.backAIC)

# backward BIC
m.backBIC <- lm(RFFT ~ Age + Education + Smoking)
summary(m.backBIC)

# forward AIC
m.forwardAIC <- lm(RFFT ~ Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + eGFR + BMI)
summary(m.forwardAIC)

# forward BIC
m.forwardBIC <- lm(RFFT ~ Education + Age + Smoking)
summary(m.forwardBIC)
```

Based on the Adjusted R-squared values, these two models have the highest values:
$R^2_{ADJ}$ =0.4622, RFFT ~ Age + Ethnicity + Education + VAT + CVD + Smoking + Hypertension + SBP + DBP + MAP + eGFR + Chol + Statin
$R^2_{ADJ}$ =0.4585, RFFT ~ Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + eGFR + BMI 

Without considering interactions, I'd probably go with the model that is more parsimonious. 


**Let's try interactions.**

Interaction AIC:
```{r}
# interaction on backward AIC
m.full.back.int.2 <- lm(RFFT ~ (Age + Ethnicity + Education + VAT + CVD + Smoking + Hypertension + 
    SBP + DBP + MAP + eGFR + Chol + Statin)^2, data=prevend)
backAIC.int.2 <- step(m.full.back.int.2, direction="backward", data=prevend)
```

Check summary of interaction on the backward AIC:
```{r}
m.backAIC.int.2 <- lm(RFFT ~ Age + Ethnicity + Education + VAT + CVD + Smoking + Hypertension + 
    SBP + DBP + MAP + eGFR + Chol + Statin + Age:Ethnicity + 
    Age:Education + Age:Smoking + Age:SBP + Age:MAP + Age:Chol + 
    Age:Statin + Ethnicity:VAT + Ethnicity:Smoking + Ethnicity:Hypertension + 
    Ethnicity:SBP + Ethnicity:DBP + Ethnicity:MAP + Ethnicity:eGFR + 
    Ethnicity:Chol + Education:Smoking + Education:Hypertension + 
    Education:SBP + VAT:SBP + CVD:Smoking + CVD:SBP + Smoking:SBP + 
    Smoking:DBP + Smoking:eGFR + Hypertension:DBP + Hypertension:MAP + 
    Hypertension:Chol + Hypertension:Statin + SBP:DBP + SBP:MAP + 
    SBP:Statin + DBP:Chol)
summary(m.backAIC.int.2)
```
The model with interaction does not look reasonable, too long.
$R^2_{ADJ}$ = 0.5176 

```{r}
# interaction on forward AIC
m.full.forw.int.2 <- lm(RFFT ~ (Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + eGFR + BMI)^2, data=prevend)
forwardAIC.int.2 <- step(m.full.intercept, scope=list(lower= ~ 1, upper= m.full.forw.int.2),
direction="forward", data=prevend)
```
result: RFFT ~ Education + Age + Smoking + Statin + Chol + Ethnicity + 
    eGFR + VAT + Education:Age + Education:Smoking + Ethnicity:eGFR + 
    Education:VAT

Check summary of interaction on the forward AIC:
```{r}
m.forwAIC.int.2 <- lm(RFFT ~ Education + Age + Smoking + Statin + Chol + Ethnicity + 
    eGFR + VAT + Education:Age + Education:Smoking + Ethnicity:eGFR + 
    Education:VAT)
summary(m.forwAIC.int.2)
```
$R^2_{ADJ}$ =0.4732. This is a little bit of improvement on $R^2_{ADJ}$. I would go with this model if we consider interaction.
 
I'm not checking numbers for interactions with BIC because BIC tends to penalize heavily on the number of parameters in general.

**Results**

Based on the above analysis, I would choose the following models:
RFFT ~ Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + eGFR + BMI with $R^2_{ADJ}$ =0.4585
RFFT ~ Education + Age + Smoking + Statin + Chol + Ethnicity + 
    eGFR + VAT + Education:Age + Education:Smoking + Ethnicity:eGFR + 
    Education:VAT, with $R^2_{ADJ}$ =0.4732.





