---
title: "Case Study Dummy Variables"
author: "Lori Hernandez"
output: word_document
---

##**Summary Statistics and Graphics**
```{r}
prevend <- read.csv("../data/Prevend_Sample.csv")
```

**Check for NA values**
```{r}
sapply(prevend, anyNA)
```

**Summary Statistics**
```{r}
summary(prevend)
```

Several columns appear to have invalid values present: VAT, which should range from 0 to 12, as negative values present (min = -1); Smoking, which should have only 0 of 1, has negative values present (min = -1); HDL, which should only be positive values, has negative values present (min = -1); and FRS, which should be only positive values, has negative values present (min = -2).  To deal with these invalid values, I will clean the data by changing all negative values to NA and then removing all rows with any NA values before starting my analysis; once I feel confident about my model and determine which predictor variables will be included, I will go back and only remove the rows with NA values in the relevant columns.

```{r}
#replacing negative values in the 4 relevant columns with NA
prevend$VAT <- replace(prevend$VAT, prevend$VAT < 0, NA)
prevend$Smoking <- replace(prevend$Smoking, prevend$Smoking < 0, NA)
prevend$HDL <- replace(prevend$HDL, prevend$HDL < 0, NA)
prevend$FRS <- replace(prevend$FRS, prevend$FRS < 0, NA)

#count the number of rows that have an NA value present
sum(!complete.cases(prevend))
```


```{r}
#remove all rows that contain any NA values
prevend <- na.omit(prevend)
```

The variable Case Number is not relevant for this analysis, so I will remove it from my data frame
```{r}
prevend <- prevend[ ,-1]
```

##**Analysis**

There are many categorical variables in this data frame; I will convert these variables to be factors
```{r}
categorical <- c("Gender", "Ethnicity", "Education", "CVD", "DM", "Smoking", "Hypertension", "Albuminuria", "Statin")
prevend[, categorical] <- lapply(prevend[, categorical], factor)
str(prevend)
```


I will look at the correlations between RFFT and the other quantitative variables
```{r}
quantitative <- c("Age", "VAT", "BMI", "SBP", "DBP", "MAP", "eGFR", "Chol", "HDL", "FRS")
round(cor(prevend$RFFT, prevend[,quantitative]), 2)
```

I am going to create a "full" model that has all of the quantitative variables as predictors
```{r}
lm.full.quant <- lm(prevend$RFFT ~ prevend$Age + prevend$VAT + prevend$BMI + prevend$SBP + prevend$DBP + prevend$MAP + prevend$eGFR + prevend$Chol + prevend$HDL + prevend$FRS)
summary(lm.full.quant)
```
Backward elimination using AIC
```{r}
mybackAIC <- step(lm.full.quant, direction="backward", data=prevend)
```
This method results in a model that uses predictors BMI, DBP, Chol, MAP, SBP, FRS, VAT, and Age; the model removed HDL and eGFR


Backward elimination using BIC
```{r}
n <-nrow(prevend)
mybackBIC <- step(lm.full.quant, direction="backward", data=prevend, k=log(n))
```

This method results in a model that uses predictors SBP, Chol, MAP, DBP, VAT, FRS, and Age; the model removed HDL, eGFR, BMI, and DBP.

Forward selection:
```{r}
my.int <- lm(RFFT~1, data=prevend)
forwardAIC <- step(my.int, scope=list(lower= ~ 1, upper= lm.full.quant), 
                   direction="forward", data=prevend)
```
This method results in a model that uses predictors Age, VAT, BMI, Chol, FRS, and SBP; the model removed MAP, DBP, eGFR, and HDL.

Forward selection using BIC:
```{r}
forwardBIC <- step(my.int, scope=list(lower= ~ 1, upper= lm.full.quant), direction="forward", data=prevend, k=log(n))
```
This method results in a model that uses predictors Age, VAT, and BMI; the model removed Chol, FRS, SBP, MAP, DBP, eGFR, and HDL.

To summarize:
Backward elimination using AIC model is RFFT ~ Age + VAT + BMI + SBP + 
  DBP + MAP + Chol + FRS with AIC of 2867.9

Backward elimination using BIC model is RFFT ~ Age + VAT + SBP + DBP + 
    MAP + Chol + FRS with AIC of 2902.57

Forward selection is RFFT ~ Age + VAT + BMI + Chol + 
   FRS + SBP with AIC of 2870.88

Forward selection using BIc is RFFT ~ Age + VAT + BMI with AIC of 2891.9

Now I will check $R^2_{ADJ}$ for each of these models
```{r}
lm.backAIC <- lm(RFFT ~ Age + VAT + BMI + SBP + DBP + MAP + Chol + FRS, data=prevend)
summary(lm.backAIC)
lm.backBIC <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS, data=prevend)
summary(lm.backBIC)
lm.forward <- lm(RFFT ~ Age + VAT + BMI + Chol + FRS + SBP, data=prevend)
summary(lm.forward)
lm.forwardBIC <- lm(RFFT ~ Age + VAT + BMI, data=prevend)
summary(lm.forwardBIC)
```
None of these models are very good, but the backward elimination using AIC model is the best of them, so I will proceed with that model.

```{r}
chosen_model <- lm(RFFT ~ Age + VAT + BMI + SBP + DBP + MAP + Chol + FRS, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```

Now it is time to consider our categorical variables. 
```{r}
par(mfrow=c(3,3))
plot(prevend$Gender, prevend$RFFT, xlab="Gender",ylab="RFFT")
plot(prevend$Ethnicity, prevend$RFFT, xlab="Ethnicity",ylab="RFFT")
plot(prevend$Education, prevend$RFFT, xlab="Education",ylab="RFFT")
plot(prevend$CVD, prevend$RFFT, xlab="CVD",ylab="RFFT")
plot(prevend$DM, prevend$RFFT, xlab="DM",ylab="RFFT")
plot(prevend$Smoking, prevend$RFFT, xlab="Smoking",ylab="RFFT")
plot(prevend$Hypertension, prevend$RFFT, xlab="Hypertension",ylab="RFFT")
plot(prevend$Albuminuria, prevend$RFFT, xlab="Albuminuria",ylab="RFFT")
plot(prevend$Statin, prevend$RFFT, xlab="Statin",ylab="RFFT")
```

We can see a clear difference in Education and Ethnicity, so I will try adding each of those to my model, one at a time
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + BMI + SBP + DBP + MAP + Chol + FRS + Education, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```
The model is better with Education, and BMI becomes less useful. I will now add Ethnicity and remove BMI

```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```
The model improves, but not by much; ethnicity doesn't seem very helpful, but I'll leave it for now.  Because we are particularly interested in Statin, I will add that next.
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity + Statin, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```

Again, the model improves slightly.  I will try adding DM now.
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity + Statin + DM, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```
The model gets worse with DM.  I will remove DM and try adding CVD.
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity + Statin + CVD, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```

The model improves very slightly.  Statin has become more significant.  I will try adding Smoking.
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity + Statin + CVD + Smoking, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```
The model improves, and FRS becomes significantly less useful.  I will try adding Hypertension
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity + Statin + CVD + Smoking + Hypertension, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```
The model improves very slightly.  I will try adding Albuminuria.
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity + Statin + CVD + Smoking + Hypertension + Albuminuria, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```

The model gets worse.  I will remove Albuminuria and try adding gender.
```{r}
chosen_model <- lm(RFFT ~ Age + VAT + SBP + DBP + MAP + Chol + FRS + Education + Ethnicity + Statin + CVD + Smoking + Hypertension + Gender, data=prevend)
summary(chosen_model)
AIC(chosen_model)
```

The model stays about the same, I will not include gender.

Out of curiosity, I will try my model building methods again but with all variables - with categorical variables as factors.
```{r}
lm.all <- lm(RFFT ~ ., data=prevend)
summary(lm.all)
AIC(lm.all)
```
Backward elimination using AIC - including categorical variables
```{r}
mybackAIC.catincl <- step(lm.all, direction="backward", data=prevend)
```

Backward elimination using BIC
```{r}
n <-nrow(prevend)
mybackBIC.catincl <- step(lm.all, direction="backward", data=prevend, k=log(n))
```

Forward selection with categorical:
```{r}
my.int <- lm(RFFT~1, data=prevend)
forwardAIC.catincl <- step(my.int, scope=list(lower= ~ 1, upper= lm.all), 
                   direction="forward", data=prevend)
```

Forward selection using BIC:
```{r}
forwardBIC.catincl <- step(my.int, scope=list(lower= ~ 1, upper= lm.all), direction="forward", data=prevend, k=log(n))
```




```{r}
#Back AIC with Categorical model
lm.backAIC.catincl <- lm(RFFT ~ Age + Ethnicity + Education + VAT + CVD + Smoking + Hypertension + 
    SBP + DBP + MAP + eGFR + Chol + Statin, data = prevend)
summary(lm.backAIC.catincl)
AIC(lm.backAIC.catincl)

#Back with BIC with Categorical model
lm.backBIC.catincl <- lm(RFFT ~ Age + Education + Smoking, data=prevend)
summary(lm.backBIC.catincl)
AIC(lm.backBIC.catincl)

#Forward selection with categorical model:
lm.forwardAIC.catincl <- lm(RFFT ~ Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + 
    BMI, data=prevend)
summary(lm.forwardAIC.catincl)
AIC(lm.forwardAIC.catincl)

#Forward selection using BIC with categorical model:
lm.forwardBIC.catincl <- lm(RFFT ~ Education + Age + Smoking, data=prevend)
summary(lm.forwardBIC.catincl)
AIC(lm.forwardBIC.catincl)
```

From everything I see, the best model is this: RFFT ~ Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + BMI  
$R^2_{ADJ}$ = 0.4559  
AIC = 4068.22

Not a great model...

```{r}
new_chosenmodel <- lm(RFFT ~ Education + Age + Smoking + Ethnicity + VAT + Chol + Statin + BMI, data=prevend)

par(mfrow=c(1,3))
plot(new_chosenmodel, c(1,2,4))
```

```{r}
rsd <- resid(new_chosenmodel)
plot(prevend$Age,rsd,xlab="Age",ylab="Residual")
abline(a=0,b=0)
plot(prevend$Ethnicity,rsd,xlab="Ethnicity",ylab="Residual")
abline(a=0,b=0)
plot(prevend$Education,rsd,xlab="Education",ylab="Residual")
abline(a=0,b=0)
plot(prevend$VAT,rsd,xlab="VAT",ylab="Residual")
abline(a=0,b=0)
plot(prevend$Smoking,rsd,xlab="Smoking",ylab="Residual")
abline(a=0,b=0)
plot(prevend$Chol,rsd,xlab="Chol",ylab="Residual")
abline(a=0,b=0)
plot(prevend$BMI,rsd,xlab="BMI",ylab="Residual")
abline(a=0,b=0)
plot(prevend$Statin,rsd,xlab="Statin",ylab="Residual")
abline(a=0,b=0)
```
Plot of observed responses vs fitted responses
```{r}
ft <- fitted(new_chosenmodel)
plot(ft,prevend$RFFT,xlab="Fitted",ylab="RFFT")
```
Not great

Variance inflation factors
```{r}
library(car)
vif(new_chosenmodel)
```

