---
title: "Case Study"
author: "Valerie Chavez"
date: "3/3/2022"
output: html_document
---

Summary Statistics and Graphics (4 points)
• Do this before formal model fitting.
• Include information on data cleaning/restructuring.

```{r}
#Data Cleaning and Restructuring

Prevend<-read.csv("Prevend_Sample.csv")

head(Prevend)
dim(Prevend)
colSums(is.na(Prevend))

#Based on this sum there appears to be no N/A's but we need to make sure the entries are valid. 

```


```{r}

range(Prevend)

#Ranges for each column to help identify if any values are out of range
#summary(Prevend)

#vat has at least one negative, #smoking has a negative, HDL and FRS have negatives 

```

Removing negatives from the data

```{r}
#replacing negatives with NAs
Prevend2<-Prevend
Prevend2[Prevend2<0]<-NA
head(Prevend2)

Prevend2<-na.omit(Prevend2)
head(Prevend2)


```


```{r}
#Summary Statistics and Graphics

#colMeans(Prevend2)
#summary(Prevend2)
#dim (Prevend2)
#We took out 42 data points

#Remove case#  from the data the box plot and regression
Prevend3<-Prevend2[-c(1)]
head(Prevend3)
boxplot(Prevend3)

```

```{r}
prevend2corset<-Prevend2[-c(1)]
corsPrevend<-cor(prevend2corset)
corrplot(corsPrevend)
#corsPrevend

```


Analysis (5 points)
• Description/discussion of analysis with enough detail that someone else could recreate your results.
• You are encouraged to try different approaches, but please restrict yourselves to the least squares
methods we have covered in this class (no other methods, such as loess or splines, should be used).
• Justify any choices that you made as part of the analysis.
• Discuss model assumptions and include diagnostic checks (plots and quantitative checks).
• As you are working toward a final model, keep in mind all that we have covered (chhecking model
assumptions, transformations, WLS, GLS, influential observations, collinearity, etc).


```{r}
#We have some variables that are measuring the same thing (3 measures of blood pressure, two measures of cholesterol)

#MAP is calculated us DBP and SBP, so it make make sense to use it, Chol total includes HDL, so it would make sense to use these. But I am going to check 


#Taking out just the quantitative variables
Prevend3_Quant<-Prevend3[-c(2,3,4,7,8,9,10,16,19)]

#head(Prevend3_Quant)


#looking at eigenvalues to detects singularity

colones<-rep(1, times=458)
x<-cbind(colones,Prevend3_Quant$SBP,Prevend3_Quant$DBP,Prevend3_Quant$MAP)
eigens_bp<-eigen(t(x)%*%x)$values

eigens_bp

cond.number<-sqrt(eigens_bp[1]/eigens_bp[4])
cond.number


#same for cholesterol
y<-cbind(colones, Prevend3_Quant$Chol, Prevend3_Quant$HDL)

eigens_chol<-eigen(t(y)%*%y)$values

eigens_chol

cond.number<-sqrt(eigens_chol[1]/eigens_chol[3])
cond.number
#not as obviously a collinearity, these might be different because, I think there is a 'give and take' with good/bad cholesterol, so we can leave both in
```


```{r}
lm.chol<-lm(RFFT~MAP+SBP+DBP,data=Prevend3)
summary(lm.chol)
anova(lm.chol)

lm2.chol<-lm(RFFT~MAP+DBP,data=Prevend3)

summary(lm2.chol)
anova(lm2.chol)

lm3.chol<-lm(RFFT~MAP,data=Prevend3)
summary(lm3.chol)
anova(lm3.chol)



```


```{r}
n<-nrow(Prevend3_Quant)
backAICchol<-step(lm.chol, arg="back",data=Prevend3_Quant)

#back BIC
backBICchol<-step(lm.chol, arg="back",data=Prevend3_Quant,k=log(n))


#both models seem to want MAP and DBP but the above R^2 weren't too great
```


I'm going to experiment with using MAP and not DBP or SBP because MAP was highly correlated with both (makes sense as they both are used to calculate MAP)

```{r}
Prevend3_Quant2<-Prevend3_Quant[-c(5,6)]

head(Prevend3_Quant2)

lm.full.quant<-lm(RFFT~., data=Prevend3_Quant2)

summary(lm.full.quant)

backAICfull<-step(lm.full.quant,arg = backwards, data=Prevend3_Quant2)



```
```{r}
#Look at any singularities 
X<-cbind(colones,Prevend3_Quant2)
eigenquant<-eigen(t(X)%*%X)
eigensquant$values
```


```{r}
lm.backaicfull<-lm(RFFT ~ Age + VAT + BMI + Chol + FRS, data=Prevend3_Quant2)

summary(lm.backaicfull)
```

```{r}
#back BIC full
backBICfull<-step(lm.full.quant,arg = backwards, data=Prevend3_Quant2,k=log(n))

```
```{r}
lm.backbicfull<-lm(RFFT ~ Age + VAT + Chol + FRS, data=Prevend3_Quant2)

summary(lm.backbicfull)
```


Adding in categoricals 
```{r}


categorical <- c("Gender", "Ethnicity", "Education", "CVD", "DM", "Smoking", "Hypertension", "Albuminuria", "Statin")
Prevend3 [, categorical] <- lapply(Prevend3[, categorical], factor)
str(Prevend3)





```

Leaving out the two other blood pressures

```{r}
PrevendQuantCat<-Prevend3[-c(12,13)]
head(PrevendQuantCat)

```

Starting with our limited model+categoricals
```{r}

#Age + VAT + BMI + Chol + FRS 




lm.limitedcat<-lm(RFFT~ Age + VAT + BMI + Chol + FRS +Gender+Ethnicity+Education+CVD+DM+Smoking+Hypertension+Albuminuria+Statin,data=PrevendQuantCat)


summary(lm.limitedcat)

```

```{r}
#Back AIC

step(lm.limitedcat,arg = backwards, data=PrevendQuantCat)


```


```{r}
summary(lm(RFFT ~ Age + VAT + Chol + FRS + Gender + Ethnicity + Education + 
    CVD + Smoking + Hypertension + Statin, data=PrevendQuantCat))
```


```{r}
step(lm.limitedcat, arg = backwards, data=PrevendQuantCat,k=log(n))
```



```{r}
summary(lm(RFFT ~ Age + Education, data=PrevendQuantCat))
```

```{r}
summary(lm(RFFT ~ Age + Education+Statin, data=PrevendQuantCat))
```



Initially in our AIC we got the quantitative variables 
Age + VAT + BMI + Chol + FRS 

Now we got either: 
formula = RFFT ~ Age + VAT + Chol + FRS + Gender + Ethnicity + 
    Education + CVD + Smoking + Hypertension + Statin
 adjusted R^2=0.4568
    or
RFFT ~ Age + Education
 adjusted R^2=0.4347
    
    
Let's try LM_full    
    
```{r}
lm.full<-lm(RFFT~.,data=PrevendQuantCat)

summary(lm.full)
```

ACI_back
```{r}
step(lm.full, arg="backwards", data=PrevendQuantCat)
```

```{r}
summary(lm(RFFT ~ Age + Gender + Ethnicity + Education + VAT + 
    CVD + Smoking + Hypertension + MAP + Chol + HDL + Statin, data=PrevendQuantCat))
```

```{r}
step(lm.full, arg="backwards", data=PrevendQuantCat,k=log(n))

```

```{r}
summary(lm(RFFT ~ Age + Education + Smoking, data=PrevendQuantCat))
```


It seems like our selection looking at the limited Quantitative variables gives a very similar selection as the full model but the full model includes Albuminuria (using AIC)

In both cased BIC picked age+education, with the additional categorical of smoking in the full model

```{r}
summary(lm(RFFT ~ Age + Education + Smoking+Statin, data=PrevendQuantCat))
```


```{r}
corqaunts<-cor(Prevend3_Quant2)

corrplot(corqaunts)

#FRS and age seem to be highly correlated, makes sense as this is a measure of congnitive function

```
```{r}

#taking out only the ones wanted in the models
Prevend4<-PrevendQuantCat[-c(8,11,13,14,16,18)]
ggpairs(Prevend4[c("Age","RFFT","Gender","Ethnicity","Education")])
pairs(Prevend4[c("Age","RFFT","Gender","Ethnicity","Education")])
#It looks like Ethnicity does not have an even spread 
```
```{r}
summary(Prevend4$Ethnicity)

#I am going to throw out ehtnicity because we have so few 1s and 2's relatively speaking 

summary(Prevend4$Education)

summary(Prevend4$Gender)
```
```{r}
ggpairs(Prevend4[c("RFFT","VAT","CVD","Smoking","Hypertension")])
pairs(Prevend4[c("RFFT","VAT","CVD","Smoking","Hypertension")])
```
```{r}
ggpairs(Prevend4[c("RFFT","MAP","Chol","Statin")])
pairs(Prevend4[c("RFFT","MAP","Chol","Statin")])
```
```{r}
Prevend5<-Prevend4[-c(3)]

lm.5<-lm(RFFT~., data=Prevend5)
summary(lm(RFFT~., data=Prevend5))
```
```{r}
backAIC<-step(lm.5, arg=backwards,data=Prevend5)

```



```{r}
par(mfrow=c(2,3))
plot(lm.5, 1:6)
```

```{r}
bc5<-boxcox(lm.5)

bc5$x[which.max(bc5$y)]
```
```{r}
#This may suggest a square root transformation
Prevend6<-Prevend5
SqrtRFFT<-sqrt(Prevend5$RFFT)
Prevend6$RFFT<-SqrtRFFT

head(Prevend5)
head(Prevend6)

colname

```


```{r}
names(Prevend6)[4]<-'sqrtRFFT'
head(Prevend6)


```

```{r}
lm.rooty<-lm(sqrtRFFT~., data=Prevend6)

summary(lm.rooty)
```


```{r}
par(mfrow=c(2,3))
plot(lm.rooty, 1:6)
```
```{r}
boxcox(lm.rooty)
```
```{r}
step(lm.rooty, arg=backwards, data=Prevend6)
```


Back aic on our transformation 
```{r}
summary(lm(sqrtRFFT ~ Age + Education + VAT + Smoking + Chol + 
    Statin,data=Prevend6))
```




